---
title: "RM Quiz3"
author: "Varun Boodram"
date: "October 28, 2014"
output: html_document
---
### Question 1
Load the cell segmentation data from the AppliedPredictiveModeling package using the commands:
```{r}
library(AppliedPredictiveModeling)
data(segmentationOriginal)
library(caret); library(rattle)
```
1. Subset the data to a training set and testing set based on the Case variable in the data set. 

```{r}
training<-segmentationOriginal[segmentationOriginal$Case=="Train",]
# note that this could also be accomplished by the followng dplyr command: 
# (https://github.com/jaeddy/practicalMlCourse/blob/master/code/quiz3.R)
# training <- segmentationOriginal %>%
#    filter(Case == "Train")
testing<-segmentationOriginal[segmentationOriginal$case=="Test",]
```

2. Set the seed to 125 and fit a CART model with the rpart method using all predictor variables and default caret settings. 

```{r, cache=TRUE}
set.seed(125)
cartModel<-train(Class ~ ., data = training, method = "rpart")
cartModel$finalModel
fancyRpartPlot(cartModel$finalModel)
```

3. In the final model what would be the final model prediction for cases with the following variable values:
a. TotalIntench2 = 23,000; FiberWidthCh1 = 10; PerimStatusCh1=2 
b. TotalIntench2 = 50,000; FiberWidthCh1 = 10;VarIntenCh4 = 100 
c. TotalIntench2 = 57,000; FiberWidthCh1 = 8;VarIntenCh4 = 100 
d. FiberWidthCh1 = 8;VarIntenCh4 = 100; PerimStatusCh1=2 

Examine each branch in the tree with the information
```{r}
newSamples <- data.frame(TotalIntench2 = c(23000, 50000, 57000, NA),
                         FiberWidthCh1 = c(10, 10, 8, 8),                         
                         PerimStatusCh1 = c(2, NA, NA, 2),
                         VarIntenCh4 = c(NA, 100, 100, 100))
rownames(newSamples)<-c("a", "b", "c", "d")
# node 1 
newSamples$TotalIntench2<45*exp(3)
# node 3
newSamples$FiberWidthCh1<9.7
```

If we follow the tree branches, 
a) WS, b) WS, c) PS, d) Not possible to predict

This is not among the possible answer choices

The correct answer is 

PS,WS, PS, Not possible (obtained by guessing; *REDO THIS QUESTION*)

### Question 2

If K is small in a K-fold cross validation is the bias in the estimate of out-of-sample (test set) accuracy smaller or bigger? If K is small is the variance in the estimate of out-of-sample (test set) accuracy smaller or bigger. Is K large or small in leave one out cross validation?

In the lecture notes, it is stated 

For k-fold cross validation

*larger k = less bias, more variance
*smaller k = more bias, less variance

so the answer to the first question is that the bias is *larger*; the answer to the second is that the variance will be *smaller*. Leave one out cross validataion leaves out a single term for comparison, and uses the rest of the data set to build the model. According to Wikipedia, "When k=n (the number of observations), the k-fold cross-validation is exactly the leave-one-out cross-validation"

### Question 3

Load the olive oil data using the commands:
```{r}
library(pgmm)
data(olive)
olive = olive[,-1]
```

These data contain information on 572 different Italian olive oils from multiple regions in Italy. Fit a classification tree where Area is the outcome variable.

```{r}
head(olive)
fit<-train(Area~., method = "rpart", data = olive)
```

Then predict the value of area for the following data frame using the tree command with all defaults

```{r}
newdata = as.data.frame(t(colMeans(olive)))
pred<-predict(object = fit, newdata = newdata)
```
What is the resulting prediction? Is the resulting prediction strange? Why or why not?
```{r}
pred
```

The result is hella strange because it is not an integer. 

# Question 4

Load the South Africa Heart Disease Data and create training and test sets with the following code:
```{r}
library(ElemStatLearn)
data(SAheart)
set.seed(8484)
train = sample(1:dim(SAheart)[1],size=dim(SAheart)[1]/2,replace=F)
trainSA = SAheart[train,]
testSA = SAheart[-train,]
```

Then set the seed to 13234 and fit a logistic regression model (method="glm", be sure to specify family="binomial") with Coronary Heart Disease (chd) as the outcome and age at onset, current alcohol consumption, obesity levels, cumulative tabacco, type-A behavior, and low density lipoprotein cholesterol as predictors.

```{r}
set.seed(13234)
fit<-train(chd~age+alcohol+obesity+tobacco+typea+ldl, method = "glm", data = SAheart, family="binomial")
```

Calculate the misclassification rate for your model using this function and a prediction on the "response" scale:

``` {r}
missClass = function(values,prediction){sum(((prediction > 0.5)*1) != values)/length(values)}

missClass(testSA$chd, predict(fit, newdata = testSA))
missClass(trainSA$chd, predict(fit, newdata = trainSA))
```


Correct answers 
Test Set Misclassification: 0.31 
Training Set: 0.27

### Question 5

Load the vowel.train and vowel.test data sets:
```{r}
library(ElemStatLearn)
data(vowel.train)
data(vowel.test) 
```
Set the variable y to be a factor variable in both the training and test set. Then set the seed to 33833.

```{r}
vowel.train$y<-as.factor(vowel.train$y)
vowel.test$y<-as.factor(vowel.test$y)
set.seed(33833)
```
Fit a random forest predictor relating the factor variable y to the remaining variables. 

```{r}
fit<-train(y~., method = "rf", data = vowel.train)
fit$finalModel
```

Read about variable importance in random forests here: http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm#ooberr The caret package uses by defualt the Gini importance. Calculate the variable importance using the varImp function in the caret package. What is the order of variable importance?

```{r}
varImp(fit)
```
Correct Answer
The order of the variables is:
x.2, x.1, x.5, x.6, x.8, x.4, x.9, x.3, x.7,x.10


What is the misclassification rate on the training set? What is the misclassification rate on the test set?

```{r}
# missClass(vowel.train$y, predict(fit, newdata = vowel.train))
# missClass(vowel.test$y, predict(fit, newdata = vowel.test))
```
